---
title: Health Data Science
output:
  html_document:
    fig_caption: yes
    theme: flatly #sandstone #spacelab #flatly
    highlight: pygments
    code_folding: show
    toc: TRUE
    toc_depth: 2
    number_sections: TRUE
    toc_float:
      smooth_scroll: FALSE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
# Only need to run these once:
#  install.packages("foreign")  
#  install.packages("tidyverse")
#  install.packages("readr")
#  install.packages("readxl")
#  install.packages("haven")
#  install.packages("survey")
#  install.packages("remotes")

# Run these every time you re-start R:
  library(foreign)
  library(tidyverse)
  library(readr)
  library(readxl)
  library(haven)
  library(survey)
  library(remotes)
  library(MEPS)
```

```{r, include=FALSE}
remotes::install_github("e-mitchell/meps_r_pkg/MEPS")
PMED21 = read_MEPS(year = 2021, type = "PMED")
```

# 1 - First Paragraph

If you need some information you can go deep in the topics here:

-   *Documentation*: <https://meps.ahrq.gov/data_stats/download_data/pufs/h229a/h229adoc.pdf>
-   *Codebook*: <https://meps.ahrq.gov/data_stats/download_data/pufs/h229a/h229acb.pdf>

```{r}
PMED21
```


**For the project i offer to consider the following topic: 'Young Adults Aged 18-24 Who Had Ever Used an Electronic Nicotine have poor general health and more likely to have asthma diagnoses'**

*Here is info about characteristics:* <https://meps.ahrq.gov/data_files/publications/st554/stat554.shtml>

*Here is questionary report:* <https://meps.ahrq.gov/survey_comp/hc_survey/paper_quest/2021/2021_SHE_SAQ.pdf>

*Codebook:* <https://meps.ahrq.gov/data_stats/download_data/pufs/h233/h233cb.pdf>

*Descriptive statistics*

```{r}
#Our data,  i put it in our holder, but if you know how to upload from github MEPS you can do it.
head(h233,9)

```

```{r}
#So our response variable is: SDENICPROD SDOH: USED ELECTRONIC NICOTINE PROD and ADSMOK42 SAQ: CURRENTLY SMOKE

range(h233$SDENICPROD)
```

